---
title: "Analysis of TRF1 SKM TMT proteomics data"
author: "Scott Lyons"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
---

## Background
Analysis of the TRF1 SKM TMT proteomics data that was searched using fragpipe and will be analyzed using MSstatsTMT.


### load packages
```{r message=FALSE}

library(tidyverse)   
library(MSstatsTMT)
library(data.table)   
library(ggprism)
library(ggrepel)
library(openxlsx)
library(pathfindR)
library(msigdbr)

```

### Load in meta data and protein data. 
Protein data will be used later on to add annotation to the final MSstats dataframe.
```{r}

#df = fread("trf1_skm_data/20220908_trf1_skm_fargpipe_msstats.csv")

protein = fread("data/20220908_trf1_skm_fargpipe_msstats_protein.tsv")

all_meta = read_csv("data/20220908_TRF1_skm_MSstats_annotation_fragpipeV1.csv", show_col_types = FALSE)


## the sample "TRF1" is a known outlier and needs to be removed from both the meta and quant data

prot_meta = all_meta %>%
  filter(!grepl('phos', Run)) %>%
  filter(BioReplicate != 'TRF1')


```

### load in msstats.csv file
currently the MSstats export from fragpipe for TMT data doesn't seem to be supposed in the proteinSummarization function so conversion first by PhilosophertoMSstatsTMTFormat is required. 
```{r}

protein_df = MSstatsTMT::PhilosophertoMSstatsTMTFormat(
  input = 'M:/Proteomics_Lab/Lab.Equipment/Proteome.Discoverer.PC/scott.lyons/fragpipe_TRF1_skm_TMT_input_phos/Run02/msstats.csv', 
  annotation = prot_meta, 
  which.proteinid = 'Protein.Accessions', 
  use_log_file = F)


```
### protein summarization 
```{r results = 'hide'}

summary = proteinSummarization(protein_df,
                               method = 'msstats',
                               reference_norm = F,
                               use_log_file = FALSE)

```
### Statistical analysis
```{r warning=FALSE, results='hide'}

levels(summary$ProteinLevelData$Condition)

compare1 = matrix(c(-1,1), nrow = 1)
comparison = rbind(compare1)
rownames(comparison) = c('TRF-AL')
colnames(comparison) = c('AL', 'TRF')

test = groupComparisonTMT(summary,
                          contrast.matrix = comparison,
                          use_log_file = F)



```


### add in mitocarta and protein annotation columns
```{r}

mitocarta = read.delim('db/Mouse.MitoCarta3.0.txt') %>%
  dplyr::select(gene_symbol = Symbol, carta_accession = accession, MitoCarta3.0_SubMitoLocalization, MitoCarta3.0_MitoPathways) %>%
  mutate(gene_symbol = gene_symbol)

add_cols = protein %>%
  dplyr::select(., `Protein ID` , `Gene`, `Protein Description`, Organism, `Protein Probability`, contains("peptide")) %>%
  distinct(., `Protein ID`, .keep_all = T)


int = test$ComparisonResult %>%
  dplyr::select(., any_of(c('Protein','Label', 'log2FC', 'pvalue', 'adj.pvalue'))) %>%
  drop_na(., pvalue) %>%
  pivot_wider(names_from = c(Label), values_from = c(log2FC, pvalue, adj.pvalue)) %>%
  separate(Protein, c('db' ,'accession', 'gene_symbol'), sep = "\\|") %>%
  left_join(., add_cols, by = c('accession' = 'Protein ID')) %>%
  left_join(., mitocarta, by = c("Gene" = "gene_symbol")) %>%
  mutate(mitocarta_gene = toupper(Gene) %in% mitocarta$gene_symbol | grepl('mitochondrial', `Protein Description`) | accession %in% mitocarta$carta_accession) %>%
  dplyr::select(., gene_symbol = Gene, accession , Description = `Protein Description`, contains('mitocarta'), contains('FC') | contains('pvalue'))

# This code block extracts the ind quant values and then takes the average of the Bioreplicates, which consists of the technical replicate data abundance values.

quant = summary$ProteinLevelData %>%
  dplyr::select(., c('Protein', 'Abundance', 'BioReplicate', 'Condition', 'TechRepMixture')) %>%
  #filter(Condition != 'pool') %>%
  group_by(Protein, BioReplicate) %>%
  mutate(mean_abundance = mean(Abundance, na.rm=T)) %>%
 # filter(TechRepMixture == 1) %>%
  dplyr::select(., -any_of(c("TechRepMixture", "Abundance", "Condition")))%>%
  #group_by(Protein) %>%
  distinct(., BioReplicate, .keep_all = T) %>%
  pivot_wider(names_from = BioReplicate, values_from = mean_abundance, names_prefix = "abundance_") %>%
  separate(Protein, c('db' ,'accession', 'gene_symbol'), sep = "\\|") %>%
  left_join(., add_cols, by = c('accession' = 'Protein ID')) %>%
  left_join(., mitocarta, by = c("Gene" = "gene_symbol")) %>%
  mutate(mitocarta_gene = toupper(Gene) %in% mitocarta$gene_symbol | grepl('mitochondrial', `Protein Description`) | accession %in% mitocarta$carta_accession) %>%
  dplyr::select(., gene_symbol = Gene, accession, Description = `Protein Description`, contains('mito', ignore.case=TRUE), contains("Protein"), contains("peptide"), contains("abundance"))


jn = left_join(quant, int)

jn2 = jn %>%
  dplyr::select(., gene_symbol, accession, Description, contains("mito", ignore.case=TRUE), contains("Protein"),
                contains("peptide"), contains("abundance"), contains("FC") | contains("pvalue"))

write.xlsx(jn, paste(Sys.Date(), 'TRF1_SKM_TMT_less_specific_carta_join.xlsx', sep='_'))


```

```{r}

quant %>%
  filter(Protein == 'contam_sp|P00761|TRYP_PIG') %>%
  group_by(Condition) %>%
  summarize(mean = mean(mean_abundance))
```



## testing
trying to figure out how MSstats deals with technical replicates
does it take the mean of each conditions within a technical replicate, then another mean of all tech replicates. Or does it take the mean of all channels first, then perfrom the group mean?

I overthought this by alot. 
```{r}

summary$ProteinLevelData %>%
  filter(Protein == "contam_sp|P00761|TRYP_PIG") %>%
  group_by(Condition) %>%
  summarize(mean = mean(Abundance))


# answer is it just takes the mean of all conditions. easy then 
```


## making spreadhheet for initial paper submission

```{r}

mitocarta2 = rename(mitocarta, accession = carta_accession)

int = map(1:8, function(x){
  
  df3 = read.xlsx("trf1_skm_data/Copy of Copy of 2022-10-19_TRF1_SKM_TMT_Fragpipe_msstats.dm.v3.xlsx", sheet = x)
  
  df3_mito = df3 %>%
    dplyr::select(., -contains("SubMito"), -contains("MitoPathways")) %>%
    left_join(., mitocarta2)
})

write.xlsx(int, paste0("SPL_", Sys.Date(), "TRF1_skm_ptx_pathways.xlsx"))

```



### volcano plot
```{r fig.width=5.0, fig.height=4.5}

volcano_plot = function(df, type, interactive, filter_mito = T, plot_title, save){
  
  mix = lapply(rownames(comparison), function(y){
    
    volcano = df %>%
      {if(filter_mito == T) filter(., mitocarta_gene == T) else .} %>%
      dplyr::select(., gene_symbol, contains(y)) %>%
      dplyr::select(gene_symbol, FC = contains("FC"), p_value = starts_with("pvalue"), q_value = starts_with("adj.pvalue"))%>%
      mutate(p_value = -log10(p_value), .keep = c("unused")) %>%
      #mutate(mitocarta = gene_symbol %in% mito_carta$Symbol | entrez_gene_id %in% mito_carta$GeneName) %>%
      drop_na(gene_symbol) %>%
      filter(!is.na(FC)) %>%
      mutate(q_sig = q_value <= 0.05)
    
    neg_chg = volcano %>%
      filter(q_value < 0.05 & FC < 0) %>%
      summarize(counts  = n())
    
    pos_chg = volcano %>%
      filter(q_value < 0.05 & FC > 0) %>%
      summarize(n())
    
    save_name = paste(y, plot_title)
    
    if(type == "protein"){
      q = ggplot(volcano, aes(x = FC, y = p_value, shape = q_sig, text = gene_symbol, q_value = q_value)) +
        geom_point(shape = 21, size = 2.5, color = 'black', stroke = 1,
                   fill = ifelse(volcano$q_value < 0.05 & volcano$FC > 0, "#A00000",
                                 ifelse(volcano$q_value < 0.05 & volcano$FC < 0, "#0000C0", "#D4D4D4")))+
        scale_x_continuous(breaks=seq(min(floor(volcano$FC), na.rm=T), max(ceiling(volcano$FC), na.rm=T), 1), 
                           limits = c(min(floor(volcano$FC), na.rm=T), max(ceiling(volcano$FC), na.rm=T)))+
        scale_y_continuous(breaks=seq(0, max(ceiling(volcano$p_value)), 1), limits = c(0, max(ceiling(volcano$p_value))))+
        theme_prism()+
        #xlim(c(-2.25,2.25))+
        geom_hline(aes(yintercept = 1.3), linetype = "dashed", size = 1.5)+
        labs(title = paste(plot_title, "\n", y, "\n", sep = " "),
             x = paste("Log2 fold change", "\n", y), y = "-log10 p-value",
             shape = "q<0.05")
      
      
      if(interactive == TRUE & save == FALSE){
        w = ggplotly(q, tooltip = 'all')
        print(w)
      }
      else if(interactive == TRUE & save == TRUE){
        w = ggplotly(q, tooltip = 'all')
        saveWidget(w, paste0(save_name, ".html"), selfcontained = T, libdir = "lib")
      }
      else if(interactive == FALSE & save == FALSE){
        q1 = q + annotate("text", label = neg_chg, x = min(floor(volcano$FC)) + 0.2, y = max(ceiling(volcano$p_value))) +
          annotate("text", label = pos_chg, x = max(ceiling(volcano$FC)) - 0.2 , y = max(ceiling(volcano$p_value)))
        
        print(q1)
      }
      else if(interactive == FALSE & save == TRUE){
        q1 = q + annotate("text", label = neg_chg, x = min(floor(volcano$FC)) + 0.2, y = max(ceiling(volcano$p_value))) +
          annotate("text", label = pos_chg, x = max(ceiling(volcano$FC)) - 0.2 , y = max(ceiling(volcano$p_value)))
        
        ggsave(q1, filename = paste0(save_name, ".pdf"), units = "in", bg='transparent', height = 5, width = 6)
        while (!is.null(dev.list()))  dev.off()
      }
    }
    
  })
  
}

v_plot = volcano_plot(df=int, type = 'protein', interactive = F, filter_mito = F,
                      plot_title = "SkM Proteomics",  save = F)    


```


### custom volcano plot
custom volcano plot to show location of proteins associated with different pathways.
```{r fig.width=5.0, fig.height=4.5}

tca = read.xlsx("trf1_skm_data/Copy of Copy of 2022-10-19_TRF1_SKM_TMT_Fragpipe_msstats.dm.xlsx", sheet = 5)[,2]
etc = read.xlsx("trf1_skm_data/Copy of Copy of 2022-10-19_TRF1_SKM_TMT_Fragpipe_msstats.dm.xlsx", sheet = 7)[,2]
atps = read.xlsx("trf1_skm_data/Copy of Copy of 2022-10-19_TRF1_SKM_TMT_Fragpipe_msstats.dm.xlsx", sheet = 6)[,2]
mr = read.xlsx("trf1_skm_data/Copy of Copy of 2022-10-19_TRF1_SKM_TMT_Fragpipe_msstats.dm.xlsx", sheet = 8)[,2]

int2 = int %>%
  mutate("tca" = accession %in% tca) %>%
  mutate("etc" = accession %in% etc) %>%
  mutate("atps" = accession %in% atps) %>%
  mutate("mr" = accession %in% mr)

to_label = c("Gls", "Mrpl16", "Slirp", "Coa6")

# Custom colored plot for TRF paper
volcano_plot = function(df, type, interactive, filter_mito = T, plot_title, save){
  
  mix = lapply(rownames(comparison), function(y){
    
    volcano = df %>%
      {if(filter_mito == T) filter(., mitocarta_gene == T) else .} %>%
      dplyr::select(., gene_symbol, where(is.logical), contains(y)) %>%
      dplyr::select(gene_symbol, where(is.logical), FC = contains("FC"), p_value = starts_with("pvalue"), q_value = starts_with("adj.pvalue")) %>%
      mutate(trans_p_value = -log10(p_value)) %>%
      drop_na(gene_symbol) %>%
      filter(!is.na(FC)) %>%
      mutate(q_sig = q_value <= 0.05)
    
    neg_chg = volcano %>%
      filter(p_value < 0.05 & FC < 0) %>%
      summarize(counts  = n())
    
    pos_chg = volcano %>%
      filter(p_value < 0.05 & FC > 0) %>%
      summarize(n())
    
    save_name = paste(y, plot_title)
    
    if(type == "protein"){
     q= ggplot(volcano, aes(x = FC, y = trans_p_value, text = gene_symbol)) +
        geom_point(data = volcano, shape = 21, size = 2.5, stroke = 1, 
                   color = ifelse(volcano$etc == TRUE, "#000080",
                               ifelse(volcano$mr == TRUE, "#A00000", "black")), 
                   fill = ifelse(volcano$etc == TRUE, "#66CCFF",
                          #ifelse(volcano$tca == TRUE, "darkorchid1",
                          #ifelse(volcano$atps == TRUE, "cadetblue1",
                          ifelse(volcano$mr == TRUE, "#FF8080",
                          ifelse(volcano$trans_p_value > 1.3, "white", "#D4D4D4"))))+
        scale_x_continuous(breaks=seq(min(floor(volcano$FC), na.rm=T), max(ceiling(volcano$FC), na.rm=T), 1), 
                           limits = c(min(floor(volcano$FC), na.rm=T), max(ceiling(volcano$FC), na.rm=T)))+
        scale_y_continuous(breaks=seq(0, max(ceiling(volcano$trans_p_value)), 1), limits = c(0, max(ceiling(volcano$trans_p_value))))+
        #geom_text_repel(data = volcano[volcano$gene_symbol %in% to_label, ], aes(label = gene_symbol))+
        theme_prism(base_size = 15)+
        geom_hline(aes(yintercept = 1.3), linetype = "dashed", size = 1.5)+
        labs(title = plot_title,
             x = paste("Log2 fold change", "\n", y), y = "-log10 p-value")
      
      
      if(interactive == TRUE & save == FALSE){
        w = ggplotly(q, tooltip = 'all')
        print(w)
      }
      else if(interactive == TRUE & save == TRUE){
        w = ggplotly(q, tooltip = 'all')
        saveWidget(w, paste(save_name, ".html"), selfcontained = T, libdir = "lib")
      }
      else if(interactive == FALSE & save == FALSE){
        q1 = q + annotate("text", label = neg_chg, x = min(floor(volcano$FC)) + 0.2, y = max(ceiling(volcano$trans_p_value))) +
          annotate("text", label = pos_chg, x = max(ceiling(volcano$FC)) - 0.2 , y = max(ceiling(volcano$trans_p_value)))
        
        print(q1)
      }
      else if(interactive == FALSE & save == TRUE){
        q1 = q + annotate("text", label = neg_chg, x = min(floor(volcano$FC)) + 0.2, y = max(ceiling(volcano$trans_p_value))) +
          annotate("text", label = pos_chg, x = max(ceiling(volcano$FC)) - 0.2 , y = max(ceiling(volcano$trans_p_value)))
        
        ggsave(q1, filename = paste0(save_name, ".pdf"), width = 6, height = 5, units = "in", bg='transparent')
        while (!is.null(dev.list()))  dev.off()
      }
    }
    
  })
  
}

v_plot = volcano_plot(df=int2, type = 'protein', interactive = F, filter_mito = T,
                      plot_title = "SkM Mito Proteins",  save = T)    


```


### pathfindR
Using the reactome gene set list from msigdb. Note that this analysis would be pretty different if an alternative gene set was used, such as kegg.
```{r message=FALSE, warning=FALSE, results='hide'}

mmu_msigdb = pathfindR::get_gene_sets_list(source = "MSigDB", species = "Mus musculus", collection = "C2", subcollection = "CP:REACTOME" )

mmu_genes = mmu_msigdb$gene_sets
mmu_desc = mmu_msigdb$descriptions

pathfind = function(df, export_name, export_data = FALSE, save_plot = FALSE){
  
  layer1 = map(rownames(comparison), function(y){
    
    int1 = df %>%
      dplyr::select(., gene_symbol, contains(y)) %>%
      dplyr::select(Gene_symbol = gene_symbol, logFC = contains('FC'), 
                    FDR_p = contains('adj.pvalue')) %>%
      drop_na(.) %>%
      as.data.frame()
    
    
    if(nrow(filter(int1, FDR_p < 0.1)) < 5){
      print("not enough proteins with a q.value < 0.1")
      
    }
    else{
      
      output = run_pathfindR(input = int1,
                             convert2alias = FALSE,
                             gene_sets = "Custom",
                             pin_name_path = "STRING",
                             #pin_name_path = path_to_pin_file,
                             custom_genes = mmu_genes,
                             custom_descriptions = mmu_desc,
                             visualize_enriched_terms = F,
                             p_val_threshold = 0.05,
                             plot_enrichment_chart = F,
                             silent_option = F,
                             n_processes = 2)
      
      
      if(export_data == TRUE){
        
        expt = list("grouped_term_output" = output,
                    "sample_term_values" = int1)
        
        write.xlsx(expt, file = paste0(Sys.Date(),"_pathfindR_results_", export_name, y, ".xlsx"))
        
      }
      
      return(output)
    }
    
  })
  
}

path_analysis = pathfind(df=int, export_name = "TRF1_SKM_MSigDB", export_data = F)



```

## creating the bubble plot
had to customize this function, originally from the pathfindR package. Needed it to change the color gradient and increase the point size.
```{r fig.width=12}

enricment_chart2 = function(result_df, plot_title, top_terms = 20, plot_by_cluster = FALSE, num_bubbles = 4, even_breaks = FALSE, save_plot){ 
  
  plt1 = map2(result_df, rownames(comparison), function(x,y){
    
    necessary <- c("Term_Description", "Fold_Enrichment", 
                   "lowest_p", "Up_regulated", "Down_regulated")
    if (!all(necessary %in% colnames(x))) {
      stop("The input data frame must have the columns:\n", 
           paste(necessary, collapse = ", "))
    }
    if (!is.logical(plot_by_cluster)) {
      stop("`plot_by_cluster` must be either TRUE or FALSE")
    }
    if (!is.numeric(top_terms) & !is.null(top_terms)) {
      stop("`top_terms` must be either numeric or NULL")
    }
    if (!is.null(top_terms)) {
      if (top_terms < 1) {
        stop("`top_terms` must be > 1")
      }
    }
    result_df <- x[order(x$lowest_p), ]
    if (!is.null(top_terms)) {
      if (plot_by_cluster & "Cluster" %in% colnames(result_df)) {
        keep_ids <- tapply(result_df$ID, result_df$Cluster, 
                           function(x) x[seq_len(min(top_terms, length(x)))])
        keep_ids <- unlist(keep_ids)
        result_df <- result_df[result_df$ID %in% keep_ids, 
        ]
      }
      else if (top_terms < nrow(result_df)) {
        result_df <- result_df[seq_len(top_terms), ]
      }
    }
    num_genes <- vapply(result_df$Up_regulated, function(x) length(unlist(strsplit(x, ", "))), 1)
    num_genes <- num_genes + vapply(result_df$Down_regulated, 
                                    function(x) length(unlist(strsplit(x, ", "))), 1)
    
    result_df$Term_Description <- factor(result_df$Term_Description, levels = rev(unique(result_df$Term_Description)))
    log_p <- -log10(result_df$lowest_p)
    
    num_genes2 = num_genes + 10
    
    result_df = cbind(result_df, total_genes = num_genes)
    
    g <- ggplot2::ggplot(result_df, ggplot2::aes_(~Fold_Enrichment, ~reorder(Term_Description, Fold_Enrichment))) +
      geom_point(ggplot2::aes(color = log_p, size = num_genes), na.rm = TRUE) +
      scale_size_continuous(range = c(2,6))+
      theme_bw()+
      ggplot2::xlab("Fold Enrichment") +
      ggplot2::labs(title = paste(plot_title, "\n", toupper(y), "\n", sep = " ")) +
      ggplot2::scale_color_gradient(low = "gray25", high = "#A00000")+
      ggplot2::labs(size = "# genes", color = expression(-log[10](p))) +
      ggplot2::theme(axis.title.y = ggplot2::element_blank(), legend.title = element_text(),
                     axis.text.y = ggplot2::element_text(color="black"),
                     axis.text.x = ggplot2::element_text(color="black"))
    
    if(save_plot == T){
      g <- ggsave(paste("bubble_plot", plot_title, ".pdf"), units = "in", bg='transparent')
      while (!is.null(dev.list()))  dev.off()
    }
    else{print(g)}
    
  }
  
  )}

plot1 = enricment_chart2(result_df = path_analysis, plot_title = "TRF1_SKM", save_plot = F)

```

### making example stacked barplot
```{r}

prot_path = read.xlsx("trf1_skm_data/Copy of Copy of 2022-10-19_TRF1_SKM_TMT_Fragpipe_msstats.dm.xlsx", sheet = 4)

paths = c("MR", "ATP", "ETC", "FA", "TCA")

prot2 = prot_path %>%
  rename(., pathway = X3) %>%
  filter(!is.na(pathway)) %>%
  filter(pathway %in% paths) %>%
  mutate(dir = ifelse(`log2FC_TRF-AL` > 0, "Up", "Down")) %>%
  mutate(dir = factor(dir, levels = c("Up", "Down"))) %>%
  group_by(pathway, dir) %>%
  summarize(n = n())

prot3 = prot_path %>%
  rename(., pathway = X3) %>%
  filter(!is.na(pathway)) %>%
  filter(pathway %in% paths) %>%
  mutate(dir = ifelse(`log2FC_TRF-AL` > 0, "Up", "Down")) %>%
  mutate(dir = factor(dir, levels = c("Down", "Up"))) %>%
  mutate(pathway = factor(pathway, levels = c("MR", "ETC", "FA", "ATP", "TCA")))
         
ggplot(prot3, aes(x = pathway, fill = dir))+
  geom_bar() +
  theme_prism(palette = 'colors') +
  labs(x="", title = "SkM mito protein pathways \n as stacked total") +
  scale_fill_manual(values = c("blue", "red"))


#write.xlsx(list(prot3, prot2), "TRF1_skm_stacked_barplot_data.xlsx", sheetName = c("cleaned_data", "final_values"))

```

